Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph for Clustering and Parameter Estimation

### 2020-08-19 summary.Rmd
updated with real data example

One popular probabilistic model for graphs with underlying communities is the 
Stochastic Block Model (SBM), which assumes that the edge probability from 
one cluster to another follows a Beronulli distribution with a fixed 
probability for each pair of communities. 
The Popularity Adjusted Block Model (PABM) was introduced by 
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the SBM to 
address the heterogeneity of edge probabilities within and between communities 
or clusters while still maintaining the community structure.

The Random Dot Product Graph (RDPG) \cite{athreya2017statistical} is another 
probabilistic model with Beronulli edge probabilities. Under this model, each 
vertex of the graph can be represented by a point in latent space such that 
the edge probability between any pair of vertices is given by their 
corresponding dot product in the latent space. The SBM is equivalent to a 
special case of the RDPG model in which all vertices in a given community share 
the same position in the latent space. It has also been shown that similar 
probabilistic graph models, (such as the Mixed Membership Stochastic Block Model
\cite{rubindelanchy2017statistical}). An analogous property exists for the 
PABM not under the RDPG model but under the *Generalized* Random Dot Product 
Graph model.

@@

**Definition 1**.
Let $G = (V, E)$ be an undirected, unweighted random graph with corresponding 
affinity matrix $A \in \{0, 1\}^{n \times n}$. Then $A$ is a random matrix 
with corresponding edge probability matrix $P$ such that 
$A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ for $i > j$ 
($A_{ji} = A_{ij}$ and $A_{ii} = 0$). Let there exist $K$ underlying 
communities in $G$, and let $n_{k}$ be the size of the $k$^th^ community in 
$G$ such that $\sum_{k=1}^K n_k = n$.

If $A$ and $P$ are organized such that $n_k \times n_l$ blocks $A^{(kl)}$ and 
$P^{(kl)}$ describe the edges and edge probabilities between communities $k$ 
and $l$, then $P^{(kl)} = \lambda^{(kl)} (\lambda^{(lk)})^\top$ for a set of 
fixed vectors $\{\lambda^{(st)}\}_{s, t = 1, ..., K}$. Each $\lambda^{(st)}$ 
for $s, t = 1, ..., K$ is a column vector of length $n_s$ (i.e., the community 
corresponding to the first index provides the vector length).

We will use the notation $A \sim PABM(\{\lambda^{(kl)}\}_K)$ to denote a random 
affinity matrix $A$ drawn from a PABM with parameters $\lambda^{(kl)}$ 
consisting of $K$ underlying clusters/communities.

@@

**Definition 2**.
Let $X \in \mathbb{R}^{n \times d}$ be latent positions of the vertices of a 
graph $G$. $X$ consists of row vectors $x_i^\top$. Let 
$A \in \{0, 1\}^{n \times n}$ be the corresponding affinity matrix.

@@

Then $G = (V, E)$ is a generalized random dot product graph with signature 
$(p, q)$ and patent positions $X$ iff its random affinity matrix can be 
described as $A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ where 
$P_{ij} = x_i^\top I_{pq} x_j$.

We will use the notation $A \sim GRDPG_{p,q}(X)$ to denote a random affinity 
matrix $A$ drawn from latent positions $X$ and signature $(p, q)$.

@@

# Real data examples

***TODO***

>>>>

Statistical analysis on graphs and networks often involve the partitioning of a 
graph into clusters. This is often motivated by the belief that there is an 
underlying and unobserved community or cluster label for each vertex of the 
graph, and edges between pairs of vertices are determined by drawing from a 
probability distribution based on the communities assigned to each pair. 
One such probabilistic model is the 
Stochastic Block Model (SBM), which assumes that the edge probability from 
one vertex to another follows a Beronulli distribution with fixed probabilities 
for each pair of community labels. 
The Popularity Adjusted Block Model (PABM) was then introduced by 
\citet*{307cbeb9b1be48299388437423d94bf1} as a generalization of the SBM to 
address the heterogeneity of edge probabilities within and between communities 
while still maintaining the community structure.

The Random Dot Product Graph (RDPG) model \cite{athreya2017statistical} is 
another graph model with Beronulli edge probabilities. Under this model, each 
vertex of the graph can be represented by a point in latent space such that 
the edge probability between any pair of vertices is given by their 
corresponding dot product in the latent space. The SBM is equivalent to a 
special case of the RDPG model in which all vertices in a given community share 
the same position in the latent space. It has also been shown that similar 
probabilistic graph models, (such as the Mixed Membership Stochastic Block Model
\cite{rubindelanchy2017statistical}). An analogous property exists for the 
PABM not under the RDPG model but under the *Generalized* Random Dot Product 
Graph (GRDPG) model.

In this paper, we will only consider undirected graphs, that is the edge value 
from vertices $i$ and $j$ is equal to the edge in the opposite direction, from 
vertices $j$ and $i$. Furthermore, we will only consider unweighted graphs with 
binary edge values. We will also assume that graphs are hollow, i.e., there 
are no self-loops.

@@

**Definition 1**. 
Let $P \in [0, 1]^{n \times n}$ be an edge probability matrix for a set of $n$ 
vertices, $V$. Each vertex has a community label $1, ..., K$, and the rows and 
columns of $P$ are arranged by community label such that $n_k \times n_l$ block 
$P^{(kl})$ describes the edge probabilities between vertices in communities 
$k$ and $l$. 
Let graph $G = (V, E)$ be an undirected, unweighted graph such 
that its corresponding adjacency matrix $A \in \{0, 1\}^{n \times n}$ is a 
realization of $Bernoulli(P)$, i.e., 
$A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ for $i > j$ 
($A_{ij} = A_{ji}$ and $A_{ii} = 0$).

If each block $P^{(kl)}$ can be written as the outer product of two vectors 
$\lambda^{(kl)} (\lambda^{(lk)})^\top$ for a set of fixed vectors 
$\{\lambda^{(st)}\}_{s, t = 1}^K$ where each $\lambda^{(st)}$ is a column vector 
of dimension $n_s$, then graph $G$ and its corresponding adjacency matrix $A$ 
is a realization of a popularity adjusted block model with parameters 
$\{\lambda^{(st)}\}_{s, t = 1}^K$.

We will use the notation $A \sim PABM(\{\lambda^{(kl)}\}_K)$ to denote a random 
adjacency matrix $A$ drawn from a PABM with parameters $\lambda^{(kl)}$ 
consisting of $K$ underlying clusters/communities.

@@

**Definition 2**.
Let $X \in \mathbb{R}^{n \times d}$ be latent positions of the vertices of a 
graph $G$. $X$ consists of row vectors $x_i^\top$. Let 
$A \in \{0, 1\}^{n \times n}$ be the corresponding adjacency matrix.

@@

Then $G = (V, E)$ is a generalized random dot product graph with signature 
$(p, q)$ and patent positions $X$ iff its random adjacency matrix can be 
described as $A_{ij} \stackrel{indep}{\sim} Bernoulli(P_{ij})$ where 
$P_{ij} = x_i^\top I_{pq} x_j$.

We will use the notation $A \sim GRDPG_{p,q}(X)$ to denote a random adjacency 
matrix $A$ drawn from latent positions $X$ and signature $(p, q)$.

@@

# Real data example

```{r}
library(mclust)
import::here(cluster.pabm, .from = '~/dev/pabm-grdpg/functions.R')

n.edges <- 20566

butterfly <- R.matlab::readMat('~/Downloads/Raw_butterfly_network.mat')

A <- butterfly$W.butterfly0
z <- butterfly$labels

keep <- c(6, 2, 9, 4)
A <- A[z %in% keep, z %in% keep]
z <- z[z %in% keep]
n <- length(z)
K <- 4

delta <- quantile(A[upper.tri(A)], 1 - n.edges / (n * (n - 1) / 2))
A <- (A > delta) * 1

z.hat <- cluster.pabm(A, K)
A.star <- A[order(z.hat), order(z.hat)]
```

For the real data example, we applied Algorithm 2 to the Leeds Butterfly 
dataset \cite{Wang_2018} consisting of visual similarity measurements among 832 
butterflies across 10 species. The graph was modified to match the example 
from \citet{noroozi2019estimation}: Only the 4 most frequent species were 
considered, and the similarities were discretized to $\{0, 1}\$ via 
thresholding. Fig. \ref{fig:butterfly} shows a sorted adjacency matrix sorted 
by the resultant the clustering.

Comparing against the ground truth species labels, Algorithm 2 achieves an 
accuracy of 63\% and an adjusted Rand index of 79\%. In comparison, 
\citet{noroozi2019estimation} achieved an adjusted Rand index of 73\%.

```{r butterfly, fig.height = 5, fig.width = 4, fig.cap = 'Adjacency matrix of the Leeds Butterfly dataset after sorting by the clustering outputted by Algorithm 2.'}
library(plot.matrix)
plot(A.star, border = NA, key = NULL, 
     xlab = '', ylab = '', main = '',
     axis.row = NULL, axis.col = NULL)
```
